\documentclass{article}

\usepackage{/home/peter/Documents/outreach/Template/packages/NUInstructions}
\usepackage{multicol}

\newcommand{\isotope}[3][]{\tikz[every node/.style = {inner sep = 0pt}]{
    \node (isotope) {#2};
    \node[left] (mass) at (isotope.north west) {\scriptsize#3};
    \node[left] at (isotope.south west) {\scriptsize#1};
    \pgfresetboundingbox
    \useasboundingbox (mass.west |- isotope.north) rectangle (isotope.east |- isotope.base);
}}

\begin{document}

\section{Introduction}

In this experiment, we'll use cloud formation to see individual particles of
alpha ($\alpha$) radiation, as well as some of the other types of radiation that
is constantly passing through our bodies.

\section{Background}

\subsection{Cloud Formation}

Water vapour exists in the air around us at all times. If there is a high enough
concentration, or the right combination of temperature and pressure, it can
condense back into liquid water. We usually see this as condensation on a
surface, such as a cold glass or a bathroom mirror. This is because having
something to condense onto (known as a nucleation site) makes the process much
more likely. Whilst we mostly see this with large surfaces, in the atmosphere
and in clouds these nucleation sites can be small particles of dust, or tiny ice
crystals. If an object has an electric charge, this also makes condensation much
more likely.

\subsection{Radioactivity}

As you will have covered (or will soon cover) at school, radioactive decay is a
completely random process during which an unstable nucleus emits a particle
(including photons) to decay into a more stable atom. The three most common
types of radiation are alpha ($\alpha$) particles, beta ($\beta$) particles, and
gamma ($\gamma$) rays. $\alpha$ particles are extremely large compared to
$\beta$ and $\gamma$ radiation, and also carry a relatively strong charge. This
means they are much more likely to interact with other particles, and of
particular interest to us much more likely to be a nucleation site for
condensation.

By placing a radioactive $\alpha$ emitter in an area of very high humidity, we
can see a trail of cloud whenever an $\alpha$ particle is released. $\alpha$
particles are absorbed by a few centimetres of air, so we should also see the
trail end after a short distance. For convenience, we'll use isopropyl alcohol
instead of water, as it has a much lower boiling point and so we can saturate
the air with vapour at room temperature.

\section{Method}

You will need:
\begin{multicols}{2}
\begin{itemize}
    \item A polystyrene block
    %
    \item A pair of gloves per person
    %
    \item A metal sheet
    %
    \item An acrylic box, with a felt lining
    %
    \item A radioactive source
    %
    \item A bottle of isopropyl alcohol (do not squeeze this bottle when picking
    it up)
    %
    \item A scoop of dry ice (don't get this yet)
    %
\end{itemize}
\end{multicols}

\task{Creating the Cloud Chamber}

\Step{} Spread a scoop of dry ice evenly over the polystyrene block.

\safety[Handling of dry ice]{Dry ice is extremely cold, and can cause burns if 
handled incorrectly. Do not touch it with bare skin, and wear gloves while 
handling it.}

\Step{} Place the metal sheet on top of the dry ice, and push down gently
(\textbf{wearing gloves}). Hold it down until it has reached the same
temperature as the dry ice. You'll be able to tell as the metal will mostly stop
making noise. \textbf{Note:} This will make a very loud noise for about a
minute. If this is likely to be a problem, ask the activity leader to step
outside while your partner does this step.

\Step{} Squirt the isopropyl alcohol onto the felt lining the acrylic box. You
do not need to tip these bottles; hold them upright, and gently squeeze to
squirt some alcohol through the nozzle. You should aim to wet the felt all the
way around, but not so much that it drips.

\safety[Use of isopropyl alcohol]{Isopropyl alcohol is flammable, and should not 
be used near an open flame. It is also toxic; do not drink it, wear gloves, and 
always wash your hands after being in any lab. Make sure you wash your hands 
before eating or drinking anything.}

\Step{} Place the radioactive source containing thorium-232 (\isotope{Th}{232}) on the centre of the
metal plate, and place the ruler perpendicular to the source rod.

\safety[Handling of radioactive \isotope{Th}{232}]{\isotope{Th}{232} is a radioactive 
substance, and should be handled with care. Wear gloves when handling it, do not 
handle it unnecessarily, and do not touch it to your skin. Always wash your 
hands after being in any lab. Make sure you wash your hands before eating or 
drinking anything.}

\Step{} Cover the radioactive source with the acrylic box. The enclosed space
should now cool and fill with vapour from the isopropyl alcohol. Watch carefully
for trails of cloud shooting out from the source.

\end{document}